---
title: "STA302_Project"
author: "Qishan (Crystal) Wei, Alina Zhang, Jieen (Jennifer) Deng, Karen Hou, Pei-Lun Hsiao"
output: html_document
date: "2026-05-19"
---

# Background

This project investigates factors associated with apartment building evaluation scores in Toronto rental properties using data from the Toronto Open Data Apartment Building Evaluation dataset.

The study focuses on how building characteristics and maintenance-related maintenance factors are associated with current apartment building evaluation scores.

# Research Question

What building characteristics and maintenance-related factors are associated with higher apartment building evaluation scores in Toronto rental properties?

# Literature Review

Previous studies suggest that housing quality and residential satisfaction are associated with building age, maintenance conditions, and structural characteristics.

Kutty (1999) found that housing adequacy is strongly associated with building characteristics and housing conditions. Zhong and Gou (2023) used RentSafeTO data to examine apartment building quality in Toronto and identified relationships between building conditions and residential quality.

# Data Description

The dataset used in this project is the Apartment Building Evaluation dataset from Toronto Open Data.\
\
The cleaned dataset contains 5304 apartment building observations after removing rows with missing values.

# Data Cleaning

The raw dataset was imported and cleaned before analysis. Relevant predictors and the response variable were selected based on findings from the literature review.

Rows containing missing values were removed using complete-case analysis. Categorical variables were converted into factor format for regression analysis.

```{r}
data <- read.csv("data/FinalProject_dataset.csv")
dim(data)
names(data)
str(data)

data_subset <- data[, c(
  "CURRENT.BUILDING.EVAL.SCORE",
  "PROPERTY.TYPE",
  "YEAR.BUILT",
  "CONFIRMED.UNITS",
  "CONFIRMED.STOREYS",
  "BUILDING.CLEANLINESS",
  "BUILDING.EXTERIOR",
  "COMMON.AREA.PESTS"
)]

data_subset[data_subset == "N/A"] <- NA
clean_data <- na.omit(data_subset)
dim(clean_data)

clean_data$PROPERTY.TYPE <- as.factor(clean_data$PROPERTY.TYPE)
clean_data$BUILDING.EXTERIOR <- as.numeric(clean_data$BUILDING.EXTERIOR)
```

```{r}
# FinalProject_dataset output
write.csv(clean_data,
          "output/cleaned_apartment_building_data.csv",
          row.names = FALSE)
```

# Linear Regression Analysis and Residual Analysis

A preliminary multiple linear regression model was fitted using predictors identified in the literature review.

Residual diagnostic plots were used to assess the assumptions of linear regression, including approximate normality and linearity.

```{r}

# Linear Regression Model 
model <- lm( CURRENT.BUILDING.EVAL.SCORE ~ PROPERTY.TYPE + YEAR.BUILT +
    CONFIRMED.UNITS + CONFIRMED.STOREYS + BUILDING.CLEANLINESS + 
    BUILDING.EXTERIOR + COMMON.AREA.PESTS,
  data = clean_data )

summary(model)

confint(model)

#
par(mfrow = c(1, 3))

#Residual vs fitted 
plot(model, which = 1) 

#QQ plot
plot(model, which = 2) 

#Standardized residual vs fitted
plot(fitted(model),
     rstandard(model),
     main = "Standardized Residual vs Fitted",
     xlab = "Fitted values",
     ylab = "Standardized residuals")

abline(h = 0, col = "red")

par(mfrow = c(1, 1))

hist(
  rstandard(model),
  main = "Histogram of Standardized Residuals",
  xlab = "Standardized Residuals",
  col = "lightgray",
  border = "white"
)


numeric_vars <- clean_data[, c(
  "YEAR.BUILT",
  "CONFIRMED.UNITS",
  "CONFIRMED.STOREYS",
  "BUILDING.CLEANLINESS",
  "BUILDING.EXTERIOR"
)]

```

# Preliminary Results

Several predictors were significantly associated with apartment building evaluation scores.

Building cleanliness, building exterior condition, and common area pests showed strong positive associations with current building evaluation scores.

The model explained approximately 51% of the variation in apartment building evaluation scores.
