---
title: "STA302_Project"
author: "Qishan (Crystal) Wei, Alina Zhang, Jieen (Jennifer) Deng, Karen Hou, Pei-Lun Hsiao"
output: html_document
date: "2026-05-19"
---

### Background

This project investigates factors associated with apartment building evaluation scores in Toronto rental properties using data from the Toronto Open Data Apartment Building Evaluation dataset.

The study focuses on how building characteristics and maintenance-related maintenance factors are associated with current apartment building evaluation scores.

### Research Question

What building characteristics and maintenance-related factors are associated with higher apartment building evaluation scores in Toronto rental properties?

### Literature Review

Previous studies suggest that housing quality and residential satisfaction are associated with building age, maintenance conditions, and structural characteristics.

Kutty (1999) found that housing adequacy is strongly associated with building characteristics and housing conditions. Zhong and Gou (2023) used RentSafeTO data to examine apartment building quality in Toronto and identified relationships between building conditions and residential quality.

### Data Description

The dataset used in this project is the Apartment Building Evaluation dataset from Toronto Open Data.\
\
The cleaned dataset contains 5304 apartment building observations after removing rows with missing values.

### Data Cleaning

The raw dataset was imported and cleaned before analysis. Relevant predictors and the response variable were selected based on findings from the literature review.

Rows containing missing values were removed using complete-case analysis. Categorical variables were converted into factor format for regression analysis.

```{r}
data <- read.csv("data/FinalProject_dataset.csv")
dim(data)

data_subset <- data[, c(
  "CURRENT.BUILDING.EVAL.SCORE",
  "PROPERTY.TYPE",
  "YEAR.BUILT",
  "CONFIRMED.UNITS",
  "CONFIRMED.STOREYS",
  "BUILDING.CLEANLINESS",
  "BUILDING.EXTERIOR",
  "COMMON.AREA.PESTS"
)]

data_subset[data_subset == "N/A"] <- NA
clean_data <- na.omit(data_subset)

clean_data$PROPERTY.TYPE <- as.factor(clean_data$PROPERTY.TYPE)
clean_data$BUILDING.EXTERIOR <- as.numeric(clean_data$BUILDING.EXTERIOR)

numeric_vars <- clean_data[, sapply(clean_data, is.numeric)]
summary(numeric_vars)
dim(clean_data)
```

```{r}
# FinalProject_dataset output
write.csv(clean_data,
          "output/cleaned_apartment_building_data.csv",
          row.names = FALSE)
```

### Data Description - Charts

Charts for describing our response variable and predictors, including a bar chart for categorical variable, histograms for other predictors and response variable, and boxplot for the categorical variable vs. other variable. 

```{r}
response_variable <-colnames(data)[1]
for (i in response_variable){
hist(x=data[[i]],xlab = i, ylab="Frequency", main=paste("Frequency vs",i))
}
predictors <-colnames(data)[3:8]
for (i in predictors){
  hist(x=data[[i]],xlab = i, ylab="Frequency", main=paste("Frequency vs",i))
}
par(mfrow=c(3,2))
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=CURRENT.BUILDING.EVAL.SCORE))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="CURRENT.BUILDING.EVAL.SCORE")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=YEAR.BUILT))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="YEAR.BUILT")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=CONFIRMED.UNITS))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="CONFIRMED.UNITS")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=CONFIRMED.STOREYS))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="CONFIRMED.STORES")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=BUILDING.CLEANLINESS))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="BUILDING.CLEANLINESS")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=BUILDING.EXTERIOR))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="BUILDING.EXTERIOR")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE),y=COMMON.AREA.PESTS))+geom_boxplot()+labs(x="PROPERTY.TYPE", y="COMMON.AREA.PESTS")
library(ggplot2)
ggplot(data, aes(x=as.factor(PROPERTY.TYPE)))+geom_bar()+labs(x="PROPERTY.TYPE", y="Frequency", title = "Bar Chart of PROPERTY.TYPE")
```

### Linear Regression Analysis and Residual Analysis

A preliminary multiple linear regression model was fitted using predictors identified in the literature review.

Residual diagnostic plots were used to assess the assumptions of linear regression, including approximate normality and linearity.

```{r}

# Linear Regression Model 
model <- lm( CURRENT.BUILDING.EVAL.SCORE ~ PROPERTY.TYPE + YEAR.BUILT +
    CONFIRMED.UNITS + CONFIRMED.STOREYS + BUILDING.CLEANLINESS + 
    BUILDING.EXTERIOR + COMMON.AREA.PESTS,
  data = clean_data )

summary(model)

confint(model)

#
par(mfrow = c(1, 3))

#Residual vs fitted 
plot(model, which = 1) 

#QQ plot
plot(model, which = 2) 

#Standardized residual vs fitted
plot(fitted(model),
     rstandard(model),
     main = "Standardized Residual vs Fitted",
     xlab = "Fitted values",
     ylab = "Standardized residuals")

abline(h = 0, col = "red")

par(mfrow = c(1, 1))

hist(
  rstandard(model),
  main = "Histogram of Standardized Residuals",
  xlab = "Standardized Residuals",
  col = "lightgray",
  border = "white"
)


numeric_vars <- clean_data[, c(
  "YEAR.BUILT",
  "CONFIRMED.UNITS",
  "CONFIRMED.STOREYS",
  "BUILDING.CLEANLINESS",
  "BUILDING.EXTERIOR"
)]

```

### Preliminary Results

Several predictors were significantly associated with apartment building evaluation scores.

Building cleanliness, building exterior condition, and common area pests showed strong positive associations with current building evaluation scores.

The model explained approximately 51% of the variation in apartment building evaluation scores.

### Model Selection

```{r}
# Check whether the response needs transformations
par(mfrow=c(2,2))
plot(model)


# Try to do the Box-Cox for Y (还不确定这个lambda图怎么解释)
par(mfrow=c(1,1))
library(MASS)
boxcox(model)

# Apply lambda = 2 for Box-Cox transformation
clean_data$Y_bc <- (clean_data$CURRENT.BUILDING.EVAL.SCORE^2 - 1)/2
model_bc <- lm(
  Y_bc ~ PROPERTY.TYPE +
    YEAR.BUILT +
    CONFIRMED.UNITS +
    CONFIRMED.STOREYS +
    BUILDING.CLEANLINESS +
    BUILDING.EXTERIOR +
    COMMON.AREA.PESTS,
  data = clean_data
)
par(mfrow=c(2,2))
plot(model_bc)

# Try to apply Log transformation
clean_data$LOG_SCORE <- log(clean_data$CURRENT.BUILDING.EVAL.SCORE)
model_log <- lm(
  LOG_SCORE ~ PROPERTY.TYPE +
    YEAR.BUILT +
    CONFIRMED.UNITS +
    CONFIRMED.STOREYS +
    BUILDING.CLEANLINESS +
    BUILDING.EXTERIOR +
    COMMON.AREA.PESTS,
  data = clean_data
)
par(mfrow=c(2,2))
plot(model_log)
```

```{r}
# Scatterplots of numerical predictors against response
# with LOWESS smoothing lines

response <- clean_data$CURRENT.BUILDING.EVAL.SCORE

par(mfrow = c(2, 3))

plot(clean_data$YEAR.BUILT,
     response,
     xlab = "Year Built",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Year Built")
lines(lowess(clean_data$YEAR.BUILT, response),
      col = "red",
      lwd = 2)

plot(clean_data$CONFIRMED.UNITS,
     response,
     xlab = "Confirmed Units",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Units")
lines(lowess(clean_data$CONFIRMED.UNITS, response),
      col = "red",
      lwd = 2)

plot(clean_data$CONFIRMED.STOREYS,
     response,
     xlab = "Confirmed Storeys",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Storeys")
lines(lowess(clean_data$CONFIRMED.STOREYS, response),
      col = "red",
      lwd = 2)

plot(clean_data$BUILDING.CLEANLINESS,
     response,
     xlab = "Building Cleanliness",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Building Cleanliness")
lines(lowess(clean_data$BUILDING.CLEANLINESS, response),
      col = "red",
      lwd = 2)

plot(clean_data$BUILDING.EXTERIOR,
     response,
     xlab = "Building Exterior",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Building Exterior")
lines(lowess(clean_data$BUILDING.EXTERIOR, response),
      col = "red",
      lwd = 2)

plot(clean_data$COMMON.AREA.PESTS,
     response,
     xlab = "Common Area Pests",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Common Area Pests")
lines(lowess(clean_data$COMMON.AREA.PESTS, response),
      col = "red",
      lwd = 2)

par(mfrow = c(1, 1))
```

```{r}
# Transformation of STOREYS
par(mfrow = c(1, 2))
clean_data$LOG_STOREYS <- log(clean_data$CONFIRMED.STOREYS)

plot(clean_data$LOG_STOREYS,
     clean_data$CURRENT.BUILDING.EVAL.SCORE,
     xlab = "Confirmed Storeys",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Storeys")
lines(lowess(clean_data$LOG_STOREYS, clean_data$CURRENT.BUILDING.EVAL.SCORE),
      col = "red",
      lwd = 2)

# Fit model with LOG_STOREYS
model_log1 <- lm(
CURRENT.BUILDING.EVAL.SCORE ~
PROPERTY.TYPE +
YEAR.BUILT +
LOG_STOREYS +
BUILDING.CLEANLINESS +
BUILDING.EXTERIOR +
COMMON.AREA.PESTS,
data = clean_data
)

summary(model_log1)

AIC(model)
AIC(model_log1)

BIC(model)
BIC(model_log1)
```

```{r}
# Transformation for UNITS
par(mfrow = c(1, 2))

clean_data$LOG_UNITS <-
log(clean_data$CONFIRMED.UNITS)

plot(clean_data$LOG_UNITS,
     clean_data$CURRENT.BUILDING.EVAL.SCORE,
     xlab = "Confirmed Units",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Units")
lines(lowess(clean_data$LOG_UNITS, clean_data$CURRENT.BUILDING.EVAL.SCORE),
      col = "red",
      lwd = 2)

# Fit model with LOG_UNITS
model_final <- lm(
CURRENT.BUILDING.EVAL.SCORE ~
PROPERTY.TYPE +
YEAR.BUILT +
LOG_UNITS +

BUILDING.CLEANLINESS +
BUILDING.EXTERIOR +
COMMON.AREA.PESTS,
data = clean_data
)

summary(model_final)

# Compare AIC/BIC
AIC(model)
AIC(model_final)

BIC(model)
BIC(model_final)

```

```{r}
par(mfrow = c(1, 2))



plot(clean_data$LOG_UNITS,
     clean_data$CURRENT.BUILDING.EVAL.SCORE,
     xlab = "Confirmed Units",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Units")
lines(lowess(clean_data$LOG_UNITS, clean_data$CURRENT.BUILDING.EVAL.SCORE),
      col = "red",
      lwd = 2)

plot(clean_data$LOG_STOREYS,
     clean_data$CURRENT.BUILDING.EVAL.SCORE,
     xlab = "Confirmed Storeys",
     ylab = "Current Building Evaluation Score",
     main = "Score vs Confirmed Storeys")
lines(lowess(clean_data$LOG_STOREYS, clean_data$CURRENT.BUILDING.EVAL.SCORE),
      col = "red",
      lwd = 2)

# Fit model with LOG_UNITS and LOG_STOREYS
model_logs <- lm(
CURRENT.BUILDING.EVAL.SCORE ~
PROPERTY.TYPE +
YEAR.BUILT +
LOG_UNITS +
LOG_STOREYS+
BUILDING.CLEANLINESS +
BUILDING.EXTERIOR +
COMMON.AREA.PESTS,
data = clean_data
)

summary(model_logs)

# Compare AIC/BIC
AIC(model)
AIC(model_logs)

BIC(model)
BIC(model_logs)

library(car)
vif(model_logs)

cor(clean_data$CONFIRMED.UNITS,
    clean_data$CONFIRMED.STOREYS)
cor(log(clean_data$CONFIRMED.UNITS),
    log(clean_data$CONFIRMED.STOREYS))
```

```{r}
# ---------------------------------------------------------
# Problematic observations: outliers, leverage, influence
# ---------------------------------------------------------

# Number of observations
n <- nrow(clean_data)

# Number of predictors in the final model, excluding intercept
p <- length(coef(model_final)) - 1


# -------------------------
# 1. Potential outliers
# -------------------------

# Studentized residuals are used to identify observations
# whose response values are poorly explained by the model.
student_resid <- rstudent(model_final)

# Since this is a large dataset, use |studentized residual| > 3
# as a practical cutoff for potential outliers.
outlier_cutoff <- 3

outlier_index <- which(abs(student_resid) > outlier_cutoff)

length(outlier_index)

# Inspect the flagged observations
outlier_table <- clean_data[outlier_index, ]
outlier_table$student_resid <- student_resid[outlier_index]

outlier_table <- outlier_table[
  order(abs(outlier_table$student_resid), decreasing = TRUE),
]

head(outlier_table, 10)
summary(outlier_table)


# -------------------------
# 2. High leverage points
# -------------------------

# Leverage values measure how unusual the predictor values are.
leverage <- hatvalues(model_final)

# Course cutoff: hii > 2(p + 1) / n
# where p is the number of predictors excluding the intercept.
leverage_cutoff <- 2 * (p + 1) / n

leverage_index <- which(leverage > leverage_cutoff)

length(leverage_index)

# Inspect high leverage observations
leverage_table <- clean_data[leverage_index, ]
leverage_table$leverage <- leverage[leverage_index]

leverage_table <- leverage_table[
  order(leverage_table$leverage, decreasing = TRUE),
]

head(leverage_table, 10)
summary(leverage_table)


# -------------------------
# 3. Influential observations
# -------------------------

# Cook's distance measures how much an observation influences
# the fitted regression model.
cooks_d <- cooks.distance(model_final)

# Common cutoff: Cook's D > 4/n
cook_cutoff <- 4 / n

cook_index <- which(cooks_d > cook_cutoff)

length(cook_index)

# Inspect influential observations
cook_table <- clean_data[cook_index, ]
cook_table$cooks_d <- cooks_d[cook_index]

cook_table <- cook_table[
  order(cook_table$cooks_d, decreasing = TRUE),
]

head(cook_table, 10)
summary(cook_table)

# Check whether any observation has extremely large influence
summary(cooks_d)
max(cooks_d)

outlier_index <- which(abs(rstudent(model_final)) > 3)
leverage_index <- which(hatvalues(model_final) > 2*(p+1)/n)
cook_index <- which(cooks.distance(model_final) > 4/n)

# Intersection = truly problematic points
problematic_index <- Reduce(intersect, list(outlier_index, leverage_index, cook_index))

length(problematic_index)

plot(hatvalues(model_final), rstudent(model_final),
     xlab="Leverage", ylab="Studentized Residuals")

abline(h=c(-3,3), col="red")
abline(v=2*(p+1)/n, col="blue")

# New Model
clean_data_no_problem <- clean_data[-problematic_index, ]
model_final_no_problem <- lm(
  CURRENT.BUILDING.EVAL.SCORE ~ PROPERTY.TYPE +
         YEAR.BUILT +
         LOG_UNITS +
         BUILDING.CLEANLINESS +
         BUILDING.EXTERIOR +
         COMMON.AREA.PESTS,
  data = clean_data_no_problem
)

# Compare ACI BCI
AIC(model)
AIC(model_final_no_problem)

BIC(model)
BIC(model_final_no_problem)

summary(model_final)

summary(model_final_no_problem)
```

```{r}
# Final model with log-transformed units
model_final <- lm(
  CURRENT.BUILDING.EVAL.SCORE ~
    PROPERTY.TYPE +
    YEAR.BUILT +
    LOG_UNITS +
    BUILDING.CLEANLINESS +
    BUILDING.EXTERIOR +
    COMMON.AREA.PESTS,
  data = clean_data
)
summary(model_final)
# Diagnostic plots for final model
par(mfrow = c(2, 2))
plot(model_final)
par(mfrow = c(1, 1))

# Actual vs fitted values
plot(fitted(model_final),
     clean_data$CURRENT.BUILDING.EVAL.SCORE,
     xlab = "Fitted Values",
     ylab = "Actual Building Evaluation Score",
     main = "Actual vs Fitted Values")

abline(0, 1, col = "red", lwd = 2)
```
